---
title: "Plotting with `ggplot2`"
subtitle: "CU Psych Scientific Computing Workshop"
tags: ["core", "R"]
weight: 400
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
```

# What Is This Tutorial? What's `ggplot2`?

ggplot2 is a package in R that allows for highly customizable and pretty plots! Here, were going to learn a few basics of making plots using `ggplot2` that will hopefully get you well on your way to making informative and beautiful data visualizations!

# Links to Files and Video Recording

The files for all tutorials can be downloaded from [the Columbia Psychology Scientific Computing GitHub page](https://github.com/cu-psych-computing/cu-psych-comp-tutorial){target="_blank"} using [these instructions](/accessing-files/){target="_blank"}. This particular file is located here: `/content/tutorials/r-core/4-plotting/index.rmd`.

For a video recording of this tutorial from the Fall 2020 workshop, please visit the <a href="/workshop-recording/session2/" target="_blank">Workshop Recording: Session 2</a> page.

# Load Packages and Data

## First load `tidyverse`, which includes `ggplot2`

We've already done this in the special first chunk up top, so it should run automatically before you run any chunk of code lower in the document.

## Load in sample data

We're going to practice here on a dataset from the 1990 NHANES (National Health and Nutrition Examination Survey). The variables are below. 

- Region -- Geographic region in the USA: Northeast (1), Midwest (2), South (3), and West (4)

- Sex -- Biological sex: Male (1), Female (2)

- Age -- Age measured in months (we'll convert this to years below)

- Urban -- Residential population density: Metropolital Area (1), Other (2)

- Weight -- Weight in pounds

- Height -- Height in inches

- BMI -- BMI, measured in kg/(m^2)

```{r}
nhanes <- read_csv(here("content", "tutorials", "r-core", "4-plotting", "NHANES1990.csv"))

nhanes <- mutate(nhanes,
                 Age = Age/12, # convert age to years
                 Urban = dplyr::recode(Urban, '1' = 'Metro Area', '2' = 'Non-Metro Area'),
                 Region = dplyr::recode(Region, '1' = 'Northeast', '2' = 'Midwest', '3' = 'South', '4' = 'West'))

nhanes
```

# Basic `ggplot2` Syntax

- `ggplot()` command usage includes saving to a variable with `a <- ggplot()`

- The general format is `ggplot(data, aes(x = [x axis variable], y = [y axis variable])` 
    - x and y variables are always specified in this `aes()` subfunction

Run this line of code:

```{r}
ggplot(nhanes, aes(x = Age, y = Weight))
```
  
  * Axes are set up the way we'd expect, and seem to have sensible values
  * Why is nothing on this graph yet? we haven't put any graphic actually on the axes yet

**We need to tell the `ggplot()` call what kind of graphic to put on the axis.**

  * A lot of the time, the syntax is `geom_[something]`

# Example Plot Types

## Scatter plots

Let's create a scatter plot using `geom_point()` first. The `ggplot()` function generates plots in layers, from "bottom" to "top", and each layer of the plot is separated by the `+` sign.

```{r}
# Although you can keep the code all on a single line, like this:
ggplot(nhanes, aes(x = Age, y = Weight)) + geom_point()

# As you add more and more layers, it can be helpful to add a
# line break after each `+` sign, like this:
ggplot(nhanes, aes(x = Age, y = Weight)) +
  geom_point()

# Note that these plots are identical! The latter format simply
# makes it easier to read the code at a glance.
```

Wow, lots of data points! Maybe we can make the points smaller to see better.

```{r}
ageWeightPlot <- ggplot(nhanes, aes(x = Age, y = Weight)) + 
  geom_point(size = .1, alpha = .3)

ageWeightPlot
```

Remember, it's a good habit to save your plots to objects, not just draw them! 

```{r}
ageWeightPlot <- ggplot(nhanes, aes(x = Age, y = Weight)) + 
  geom_point(size = 1, alpha = .2, aes(color = Region, pch = Region))

ageWeightPlot
```

## Histograms

```{r}
weightHistogram <- ggplot(nhanes) +
  geom_histogram(aes(x = Weight), bins = 100, fill = 'purple')

weightHistogram
```

## Dot plots organized by grouping factor

Question: what if we want to look at distribution of weights by region in the nhanes data?

With `ggplot2`, if x is a factor (discrete, not continuous) we can plot dots as a function of the factor.

```{r}
heightByUrban <- ggplot(nhanes, aes(x = Urban, y = Height)) + 
  geom_point()

heightByUrban
```

Whoa! So much data, it's hard to see anything. Let's use `geom_jitter()` with size = .05, width = .1 instead.

```{r}
heightByUrban <- ggplot(nhanes, aes(x = Urban, y = Height)) +
  geom_jitter(size = .05, width = .1, height = 0, alpha = .1)

heightByUrban
```

## Violin plots to visualize distributions

Still a TON of data! Here's a great tool I really like for plotting the density of distributions of points like this!

```{r, warning=FALSE}
# We can use 'source' to pull bits of code from github, as well as local files
source("https://gist.githubusercontent.com/benmarwick/2a1bb0133ff568cbe28d/raw/fb53bd97121f7f9ce947837ef1a4c65a73bffb3f/geom_flat_violin.R") 

heightByUrban +
  geom_flat_violin(position = position_nudge(x = .15, y = 0), alpha = .7)
```

Now, we can really begin to see the skewedness of these weight distributions.

## Summary plots

Plotting data points is all well and good, but what if we want to use our plots to summarize distributions? We'll do that here.

```{r}
weightByRegion <- ggplot(nhanes, aes(x = Region, y = Weight)) + 
  stat_summary(fun = "mean", geom = "point")

weightByRegion +
  ylim(0,400)

# Here is how to plot the group means with confidence intervals
weightByRegionConf <- ggplot(nhanes, aes(x = Region, y = Weight)) + 
  stat_summary(fun.data = "mean_cl_boot", fun.args = list(conf.int = .95))

weightByRegionConf
```

It might be useful to look at the summary with the raw data too...just to put any potential region differences in context.

Be sure to note that when we do this, `ggplot()` adds these "layers" to the plot from the bottom up!

```{r}
# If we just add the `geom_jitter()` call to the end of the function,
# the raw data is the last layer added to the plot, which obscures the means!
rawPlusSummary <- ggplot(nhanes, aes(x = Region, y = Weight)) +
  stat_summary(fun.data = "mean_cl_boot", fun.args = list(conf.int = .95), color = 'red') +
  geom_jitter(size = .1, alpha = .1, height = 0, width = .1)

rawPlusSummary

# In this version, we have reversed the ordering of `geom_jitter()` and
# `stat_summary()` so we can easily see the raw data AND the means
rawPlusSummary <- ggplot(nhanes, aes(x = Region, y = Weight)) + 
  geom_jitter(size = .1, alpha = .1, height = 0, width = .1) +
  stat_summary(fun.data = "mean_cl_boot", fun.args = list(conf.int = .95), color = 'red')

rawPlusSummary
```

We can also add the density plots!

```{r}
rawPlusSummary <- ggplot(nhanes, aes(x = Region, y = Weight)) + 
  geom_jitter(size = .1, alpha = .1, height = 0, width = .1) +
  geom_flat_violin(position = position_nudge(x = .15, y = 0)) +
  stat_summary(fun.data = "mean_cl_boot", fun.args = list(conf.int = .95),
               color = 'red', position = position_nudge(x = .2, y = 0))

rawPlusSummary
```

Now that we can visualize the distributions, we can see that weight is bimodally distributed (probably because of the young children in this dataset).

## Heatmaps

Sometimes we might want to make a heatmap to look at the value of variable based on a 2D grid of two other variables.

This is kind of a silly example, but say we wanted to map out the number of observations in our dataset as a factor of region and neighborhood type.

```{r}
# group the data
nhanesGroup <- nhanes %>%
  group_by(Region, Urban) %>%
  summarize(Observations = n())

ggplot(nhanesGroup, aes(x = Urban, y = Region)) + 
  geom_tile(aes(fill = Observations)) +
  scale_fill_gradient(low = "blue", high = "red") + 
  theme_bw()
```

## Time series plots

**Working with a different dataset here**

Let's make up some very simple data on the prices of two different items from 1978-2017.

```{r}
years <- 1978:2017
item1<- rnorm(40,100,5)
item2 <- 1:40 + rnorm(40,100,5)

# Helpful to put it in long form using pivot_longer
timeFrame <- tibble(years = years, item1 = item1, item2 = item2) %>%
  pivot_longer(cols = starts_with('item'), names_to = 'item', values_to = 'price') %>%
  mutate(se = runif(n(), 1,5))
```

We can plot the time series using `geom_point()` and connect the times using `geom_line()`, coloring by item.

```{r}
ggplot(timeFrame, aes(x = years, y = price, color = item)) +
  geom_point() +
  geom_line(aes(lty = item)) +
  theme_bw()
```

Now, let's get some representation of our uncertainty into the plot! Notice that there is an included 'se' column for the standard error of each observation. Let's plot error bars of +/- 1 standard error above and below each point.

  * We can specificy the range of the errorbars with the `ymin` and `ymax` arguments
  * It can also look nice to set width to 0

```{r}
ggplot(timeFrame, aes(x = years, y = price, color = item)) +
  geom_point() +
  geom_line() +
  geom_errorbar(aes(ymin = price - se, ymax = price + se), width = 0) +
  theme_bw() +
  labs(x = 'Year', y = 'Price', color = 'Item')
```

Alternatively, we can use shading to express our uncertainty more continuously. This time, let's shade the error within **2 standard errors** of each measured point.

```{r}
ggplot(timeFrame, aes(x = years, y = price)) +
  geom_point(aes(color = item)) +
  geom_line(aes(color = item)) +
  geom_ribbon(aes(ymin = price - 2*se, ymax = price + 2*se, fill = item), 
              alpha = .2, show.legend = F) +
  theme_bw() +
  labs(x = 'Year', y = 'Price', color = 'Item') +
  scale_color_brewer(palette = 'Dark2') +
  scale_fill_brewer(palette = 'Dark2')
```

Notice we've had to reformat a few calls here to adjust the aesthetic mapping...this happens sometimes when we want certain mappings to apply to ONLY certain parts of the plot. When we put the `aes()` call inside a `geom()` call, the mapping applies only to that geometrical object.

*Remember! It's very important to always display the predictive uncertainty along with the estimates or mean predicted by your model. Otherwise, we don't have any idea of how confident the model's predictions are.*

# Fitting Lines to the Data

Let's say we think there might be a linear relationship between height and weight.

We can use `geom_smooth` for this and `method = 'lm'` specifically for a linear model
Also `level = .95` can specify the 95% confidence interval about the estimate at each x value.

```{r}
heightByWeight <- ggplot(nhanes, aes(x = Height, y = Weight)) + 
  geom_point(size = .1, alpha = .1) + 
  stat_smooth(method = 'lm')

heightByWeight
```

Hmm, this actually looks like it's giving us some pretty bad predictions. We're not going to get into the stats of this now, but we can also plot using `auto` which is a mix of models, and might be a bit smarter.

```{r}
heightByWeight <- ggplot(nhanes, aes(x = Height, y = Weight)) + 
  geom_point() + 
  stat_smooth(method = 'auto', level = .99, alpha = 1, color = 'green', fill = 'red')
heightByWeight
```

# Plot Formatting

## Titles and axis labels

The `labs()` command can be added to `ggplot2` with different arguments, lke `x`, `y`, or `title` to make the plots clearer.

```{r}
heightByUrban +
  labs(x = 'Neighborhood', y = 'Height in Inches', title = 'Height by Urban Status')
```

Note, if we want to change labels for _factors_, not just axes, it's easier to do that using `tidyverse`.

## Facetting

It's useful to have several plots in a panel sometimes, rather than just one. 

So for this data set, say we want to plot relationships between height and weight, but by region.

We can do this with `facet_wrap('Region')`.

```{r}
facetPlot <- ggplot(nhanes, aes(x = Height, y = Weight)) + 
  geom_point(alpha = .2, size = .1) + 
  stat_smooth() + 
  facet_wrap('Region')

facetPlot
```

We can even do multiple factors.

```{r}
multiFacet <- ggplot(nhanes, aes(x = Height, y = Weight)) + 
  geom_point(alpha = .2, size = .5, aes(color = Urban)) + 
  stat_smooth() + 
  facet_grid(Region ~ Urban, scales = 'free_y')

multiFacet
```

## Color

We can color either continuously or discretely, depending on how a variable is represented in R.

We put `color = Height` into the `aes()` function because it's a grouping factor.

**Continous Example:**

```{r}
weightByRegion <- ggplot(nhanes, aes(x = Region, y = Weight, color = Height)) + 
  geom_jitter(width = .1) + 
  labs(x = 'Region of US', y = 'Weight (lbs)', title = 'Weight by Region')

weightByRegion
```

**Discrete Example:**

```{r}
discretePlot <- ggplot(nhanes, aes(x = Height, y = Weight, color = Region)) +
  geom_point() 

discretePlot
```

It is easy to choose your own custom colors (as well as using R presets), but we're not going to get into that right at this moment.

## Themes

We can use themes to make our plots prettier, and also to customize the gridlines. 

```{r}
discretePlot + 
  theme_bw()

discretePlot + 
  theme_minimal()

discretePlot + 
  theme_void()

discretePlot + 
  theme_classic()

facetPlot + 
  theme_bw() + 
  labs(title = 'Weight by Height and Region')
```

# Saving to Files

`ggsave()` includes arguments for `filename`, `plot`, `dpi`, `width`, and `height` (among other things).

We can use a variety of file formats.

```{r, eval = FALSE}
ggsave('newPlotTest.pdf', plot = facetPlot, dpi = 300, width = 5, height = 5)
```

# Final Points

Basic `ggplot2` will get you a LONG way.

Also, there is much more `ggplot2` can do for making your plots very pretty, and also plotting lots of complex models.

Unlike Excel and SPSS, which can often be cranky and difficult to bend to your will in customizing plots, `ggplot2` is really easy to work with to make your graph look the way you want.

**Next:** [Modeling (Basic modeling in R)](/tutorials/r-core/5-modeling/)
