---
title: "Descriptives_Data_Challenge"
output: html_document
---

# Data challenge for descriptive statistics


0. We're going to continue working with the Uncapher et al. (2016) dataset for this challenge! First, load the *uncapher_2016_one_condition_dataset.csv* file into a dataframe 

```{r}

```


1. Make a histogram of the variable called "Dprime". This variable represents (normed) hit rate minus (normed) false alarm rate during the working memory task. 

```{r}

```



2. What is the skew and kurtosis of Dprime? Is it normally distributed? 
```{r}
# hint: there is missing data in the dprime variable. how do we get around that?



```

3. Make a boxplot comparing Dprime between high multitaskers (groupStatus: HMM) and low multitaskers (groupStatus: LMM). 
```{r}


```

4. what is the mean dprime score for high multitaskers and low multitaskers?
```{r}



```

5. Make a plot showing the relationship between multitasking and dprime.
Is this relationship positive or negative?  


```{r}


```

