---
title: "Descriptives_Data_Challenge"
output: html_document
---

# Data challenge for descriptive statistics
```{r}

library(car)
library(moments)
library(tidyr)
library(dplyr)

```

Using the same dataset, complete the following exercises. 
```{r}
# this dataset examines the relationship between multitasking and working memory.
# original paper by Uncapher et al. 2016:  https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4733435/pdf/nihms712443.pdf

# open the data from csv 
df <- read.csv("uncapher_2016_one_condition_dataset.csv")
```

1. Make a histogram of the variable called "Dprime". Add colors and labels. This variable represents (normed) hit rate /(normed) false alarm rate during the working memory task. 

```{r}

hist(df$dprime, col = "blue", breaks = 20)


```



2. What is the skew and kurtosis of Dprime? Is it normally distributed? 
```{r}

skewness(df$dprime, na.rm = TRUE)

kurtosis(df$dprime, na.rm = TRUE)

shapiro.test(df$dprime)
```

3.Make a boxplot comparing Dprime between high multitaskers (groupStatus: HMM) and low multitaskers (groupStatus: LMM). 
```{r}

boxplot(df$dprime, df$groupStatus)

```

4. what is the mean dprime score for high multitaskers and low multitaskers?
```{r}

grouped <- group_by(dprime, groupStatus)

# now we will extract summary statistics from the grouped data
summarise(grouped, mean=mean(dprime), sd = sd(dprime),  min=min(dprime), max=max(dprime)))


```

5. Make a plot showing the relationship between multitasking and dprime.
Is this relationship positive or negative?  

```{r}

plot(df$dprime ~ df$mmi) # negative. 


```

