---
title: "R TUTORIAL: Descriptives"
output: 
  html_document: 
    toc: true
    toc_float: true
---

# Goals for this tutorial:
1. Learn basics of plotting descriptives in R 
2. Learn tools to generate descriptive statistics in R 


#install and load packages that you'll need
```{r setup, include=FALSE}


#install.packages("car")
#Companion to Applied Regression
#Functions and Datasets to Accompany J. Fox and S. Weisberg, An R Companion to Applied Regression, Second Edition, Sage, 2011.
# https://cran.r-project.org/web/packages/car/car.pdf

#install.packages("moments") 
# useful descriptive calculationss
#https://cran.r-project.org/web/packages/moments/moments.pdf


library(car)
library(moments)
library(tidyr)
library(dplyr)
```

# Let's explore a sample dataset! 
```{r}
# this dataset examines the relationship between multitasking and working memory.
# original paper by Uncapher et al. 2016:  https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4733435/pdf/nihms712443.pdf

# open the data from csv 
df <- read.csv("uncapher_2016_one_condition_dataset.csv")

# check the data format
head(df)

# check the variable names 
names(df)
```

First we will check the data to make sure 
(a) it's a dataframe
(b) what the column names are 
(c) how many rows and columns 
(d) snapshot of the data itself
```{r}
# check that it's a dataframe 
class(df) 
# list out column names (Variable names)
names(df) 

# how many rows ?
nrow(df) 

# how many columns? 
ncol(df) # number of columns 

# get a snaphot of the data inside the dataframe
str(df)  

# how many subjects are in this dataframe?
Nsubs <- length(unique(df$subjNum))

```

# 1.A. graphing skew/ normality of variables 
plot histogram of continuous variables
```{r}
# plot histogram of media multitasking
hist(df$mmi)
# what is the skew?

# plot histogram of ADHD measure
hist(df$adhd)

# r picks the bins it thinks worst best, but sometimes you want to change them. let's adjust the breaks so there are no spaces in the histogram.
hist(df$adhd, breaks = 8)

# plot histogram of impulsivity measure
hist(df$bis)
```

Next, let's make the histogram prettier with some extra arguments.
base R graphics have many more options, these are just a few to get you started.
```{r}
# add light blue color using col = "light blue"
# add a yaxis label and an x axis label using ylab and xlab

hist(df$mmi, col= "light blue", main = "", ylab = "Frequency", xlab = "Media Multitasking")
```


#1.B. Quantative assessments of skew:
```{r}
#calculate skew for mmi. is it positive or negative? 

skewness(df$mmi)

#calculate kurtosis for mmi. is it positive or negative? 
kurtosis(df$mni) 
# hint: positive kurtosis means that the tails are thinner than normal distribution.

#Use the shapiro test, which tests null hypothesis that the samples came from a normal distribution.
shapiro.test(df$mmi)  
```
 

# 2. Basic Descriptives
let's learn how to get range, mean, sd, minimum, maximum, etc.
```{r}
# provide the range of mmi variable
range(df$mmi)

# average score  of mmi 
mean(df$mmi, na.rm = TRUE)

# standard deviation of mmi
sd(df$mmi, na.rm = TRUE)

# the simplest way to get summary descriptives on a variable 
summary(df$mmi)

 
# we can also combine them into a table, using the summarise function 
my_table <- summarise(df, mean=mean(mmi), stdev = sd(mmi),  min=min(mmi), max=max(mmi))

# if you want to save this table for later, write to a csv ! 
write.csv(my_table, "multitasking_summary_table.csv" )


#notes: it will automatically save to your working directory unless you specify otherise 
# remember to check your working directory, use gwd() 
# to change your working directory, use setwd() 
# or, add the path manually "/dir/dir/dir/filename.csv"
```

Quantiles 
```{r}

#The GroupStatus variable represents the 1st and 4th quartiles of media multitasking score (variable: mmi, LMM and HMM)

# Let's extract the mmi value at the 1st and 4th quartiles 
quantile(df$mmi, (0.25), na.rm = TRUE)

quantile(df$mmi, (0.75), na.rm = TRUE)

```

# 2B. Descriptives with grouping
We can also generate descriptive statistics based on grouping variables. 
Here we will look at impulsitivity (bis) by group: high or low media multitasking (HMM or LMM)

```{r}
# the group_by function is from our friend, from tidyverse.
library (tidyr)
# the first argument is your variable of interest (mmi)
# second & third arguments are your grouping variables (factors)
# group by the between subject variable: groupStatus (high, low media multitasking))
grouped <- group_by(df$bis, df$groupStatus)

# now we will extract summary statistics from the grouped data
summarise(grouped, mean=mean(bis), sd = sd(bis),  min=min(bis), max=max(bis))

```

# 3.Boxplots 
We will use the basic boxplot function available in base R. Boxplots can be created for individual variables quite easily and give you a quick idea of the spread of your data. Here, we will make a boxplot for self-reported media multitasking (mmi).

```{r}

boxplot(df$mmi)

# outliers are automatically labeled as dots.
# you can exclude them if you want with a simple argument 
# notice that the y axis also automatically changes!
boxplot(df$mmi, outline = F)

# what if you prefer horizonal plots? 
boxplot(df$mmi, horizontal = TRUE)
```

#3B. Boxplots with grouping 
We can also use grouping variables in boxplots. Here we will plot ADHD symptoms by group: high or low media multitasking (HMM or LMM)
```{r}
#mmi = y variable, groupStatus is our (x) variable
# we are only interested in the high multitasking and and low multitasking groups, so we will filter the data to remove those in the middle (and therefore not labeled for groupStatus)

df.group <- subset(df, groupStatus != "")
df.group$groupStatus <- as.factor(df.group$groupStatus)
boxplot(df$bis ~  df$groupStatus, data=df)
#*** PAUL HOW TO GET RID OF 3rd empty LEVEL?
  
# let's make this one prettier by adding axis labels and colors 
boxplot(adhd ~ groupStatus, data=df.group,
  	xlab="Multitasking Group", ylab="ADHD", col = c("light blue", "light green"))
```

# 5. Relationships between two continuous variables

```{r}
# this is the most basic plot you can make. 
#don't worry, later we will show you how to make much prettier graphs! 

# graph the relationship between ADHD (x) and impulsivity (y)
plot (df$adhd, df$bis)
```

# 6. Relationships of many variables
We can use scatterplot matrices  which are a handy & quick way of visualizing relationships between multiple variables in a dataset. This works best when you have 3-5 variables (it gets too hard to read after that).
Let's look at the relationships between multitasking, impulsitivity, and adhd symptoms.

```{r}
ncol(df)

# We will use the pairs function 
# in order to isolate the 3 variables we want, indexing [16:18] variables 
pairs (df[16:18])
# There are positive relationships evident between each pair of variables! 
# We will explore these relationshisp more in the statistics tutorial.
```

