---
title: "Data Manipulation"
subtitle: "CU Psych Scientific Computing Workshop"
tags: ["core", "R"]
weight: 300
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
```

# Goals of This Lesson

1. Learn how to group and categorize data in R 
2. Learn how to generate descriptive statistics in R

# Links to Files and Video Recording

The files for all tutorials can be downloaded from [the Columbia Psychology Scientific Computing GitHub page](https://github.com/cu-psych-computing/cu-psych-comp-tutorial){target="_blank"} using [these instructions](/accessing-files/){target="_blank"}. This particular file is located here: `/content/tutorials/r-core/3-datamanipulation/index.rmd`.

For a video recording of this tutorial from the Fall 2020 workshop, please visit the <a href="/workshop-recording/session2/" target="_blank">Workshop Recording: Session 2</a> page.

# Load Data

This dataset examines the relationship between multitasking and working memory ([link to the original paper by Uncapher et al., 2016](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4733435/pdf/nihms712443.pdf){target="_blank"}).

```{r, message = FALSE, warning = FALSE}
# use read.csv function to open data into R 
df <- read_csv(here("content", "tutorials", "r-core", "3-datamanipulation", "uncapher_2016_repeated_measures_dataset.csv"))

# we will continue using tidyverse tools
# so we have library()-ed tidyverse at the very top of this document
```

# Familiarizing Yourself with the Data

Quick review from [Data Cleaning](/tutorials/r-core/2-datacleaning/): take a look at the basic data structure, number of rows and columns.

```{r}
# base R functions
df
nrow(df)
ncol(df)
names(df)
```

# Selecting Relevant Variables

Sometimes datasets have many variables that are unnecessary for a given analysis. To simplify your life, and your code, we can select only the given variables we'd like to use for now.

```{r}
# tidyverse select() function 
df <- df %>%
  select(subjNum, groupStatus, adhd, hitRate, faRate, dprime)
```

# Basic Descriptive Statistics

## Summarizing data

Let's learn how to make simple tables of summary statistics.
First, we will calculate summary info across all data using `summarize()`, a useful function for creating summaries. Like `mutate()`, it can take mutiple functions as arguments. Note that we're not creating a new object for this summary (i.e. not using the `<-` symbol), so this will print but not save.

```{r}
df %>% 
  summarize(min_HR = min(hitRate, na.rm = T),
            max_HR = max(hitRate, na.rm = T), 
            mean_HR = mean(hitRate, na.rm= T), 
            sd_HR = sd(hitRate, na.rm = T))
```

# Grouping data

Next, we will learn how to group data based on certain variables of interest.

We will use the `group_by()` function in tidyverse, which will automatically group any subsequent actions called on the data. 

```{r}
# Split the summary statistics by group
meanSummary <- df %>% 
  group_by(groupStatus) %>%
  summarize(meanHitRate = mean(hitRate, na.rm= T), 
            meanFalseAlarm= mean(faRate, na.rm= T),
            meanDprime = mean(dprime, na.rm = T))
```

We can group data by more than one factor. Let's say we're interested in how levels of ADHD interact with `groupStatus` (multitasking: high or low). We will make a factor for ADHD (mean-split), and add it as a grouping variable.

```{r}
# If adhd score is lower than mean, label "low", else label "high"
df <- df %>% 
   mutate(adhdF = if_else(adhd < mean(adhd), "Low", "High")) 

table(df$adhdF)
```

Then we'll check how evenly split these groups are.

```{r}
# How many data points are in each group (adhdF x groupStatus)?
# count() will calculate this for us.
df %>%
  count(adhdF, groupStatus)
```

Then we'll calculate some summary info on these groups.

```{r}
df %>%
  group_by(adhdF, groupStatus) %>%
  summarize(meanHitRate = mean(hitRate, na.rm= T), 
            meanFalseAlarm= mean(faRate, na.rm= T),
            meanDprime = mean(dprime, na.rm = T))
```

# Extra: Working with a Long Dataset 

This is a repeated measures ("long") dataset, with multiple rows per subject. This makes things a bit tricker, but we are going to show you some tools for how to work with "long" datasets.

## Counting unique subjects

```{r}
# Get a list of subjects using unique()
SubList <- unique(df$subjNum)

SubList

# how many subjects are in this dataframe?
Nsubs <- length(SubList)
```

## Calculating number of trials per subject

```{r}
df %>%
  count(subjNum)
```

## Combining summary statistics with the full dataframe

For some analyses, you might want to add a higher level variable (e.g. subject average hitRate) alongside your long data. We can do this by using mutate instead of summarize. Note: you'll have repeating values for the average column.

```{r}
avgHR <- df %>% 
  group_by(subjNum) %>% 
  mutate(avgHR = mean(hitRate, na.rm=T))

# You should now have an avgHR column in df, which will
# repeat within each subject, but vary across subjects.
```

# Saving Your Work

## Saving tables into .csv files

```{r}
basic_descriptives <-  df %>% 
  # if adhd score is lower than mean, label "low", else label "high""
  mutate(adhdF = if_else(adhd < mean(adhd), "Low", "High")) %>%
  # split the summary statistics by group 
  group_by(adhdF, groupStatus) %>%
  summarize(meanHitRate = mean(hitRate, na.rm= T), 
            meanFalseAlarm= mean(faRate, na.rm= T),
            meanDprime = mean(dprime, na.rm = T))
```

```{r, eval = FALSE}
# To save this out, use write_csv() to save a CSV file (which can open in
# other programs like Excel) or save() to save a combined file comprising
# the full data and descriptives (which can only open in R/RStudio).
write_csv(basic_descriptives, path = here("content", "tutorials", "r-core", "3-datamanipulation", "myDescriptives.csv"))
save(basic_descriptives, df, file = here("content", "tutorials",, "r-core", "3-datamanipulation", "StudyData.rda"))

# Note: These files will automatically save to your working directory
# unless you specify otherwise. To do so, use here() to indicate the
# location where you would like to save the file.
```

**Next:** [Plotting (Graphing your data using `ggplot2`)](/tutorials/r-core/4-plotting/)
