---
title: "Programming Challenge"
author: "Monica Thieu"
date: "November 26, 2017"
output: html_document
---

```{r setup, include=FALSE}
# Don't worry about what's in here
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# In this chunk:
# In the below code chunk, initialize a variable and set its value to any integer of your choice.
# Initialize another variable, which will "check" the value of your first integer variable using some code, and have this variable's value be TRUE if your integer is an ODD number, and FALSE if your integer is an EVEN number.
# Test your code using both positive and negative as well as both odd and even numbers, and zero.
```

Here's a dataframe containing a bit of information about Shakespeare's plays. Let's find out some information about this! The answer to each question should be expressed as the output of a command or a series of commands. "I looked and this is what the answer was" does not count!

```{r}
load("programming-challenge.data.rda")
```

How many total plays did Shakespeare write?
```{r}

```

What genres of plays did he write?
How many different genres of plays did he write, according to this data?
```{r}

```

How many plays of each genre did he write? Be able to express this both as the number of plays, and the proportion of his total plays.
```{r}

```

On average, how many words were in a Shakespeare play?

Try rounding this to the nearest word! If you don't know what function to use, you can look it up :)
```{r}

```

How many words were in each genre of play? Which genre has the greatest mean number of words per play?
```{r}

```

How many plays have fewer than twenty thousand words?
```{r}

```

Show the names of all plays that are either comedies or tragedies.
```{r}

```

As you can see, this list is sorted alphabetically. What are the 5th through 8th plays alphabetically?
```{r}

```

Create a new column on the dataframe shakes where all the titles are in all caps. Name the column something sensible.
```{r}

```

Write code that will do the following.
If you give the code the name of a play, written as a string, it should do the following:

- If the play is a **comedy,** output the number of words in the play
- If the play is a **tragedy,** output the index of the play's row in the dataframe
- If the play is a **history,** output the number of characters in the play's name (spaces count)

Bonus: Can you make your code case-insensitive to the play name that's fed into the code? (assume it's still spelled correctly)
```{r}

```

