---
title: "Programming Lesson Plan"
author: "Monica Thieu"
date: "November 26, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Intro

Programming is telling a computer to do some action to some input info to get output info that you want. We will learn about this info that goes in and out, as well as the actions that transform info into what we want.

## Working directories, file paths, etc

Before we actually get started with messing in R, we need to understand how R interacts with files on your computer. Any data that gets loaded in or saved out is saved on your computer proper, and you can access all of these files through R.

```{r explain working directories}
install.packages("knitr")
getwd()
```

```{r explain file paths}
# Absolute, e.g. the following
"/Users/mthieu/Dropbox"
# Relative
"mthieu/Dropbox"
# Note that absolute file paths will typically START with a forward-slash / character, while relative paths will NEVER start with / . Why is this the case?
# When would you want to use either one?

"../" # This allows you to go up one level, so backwards in your file directory 
```


# WHAT

## Variables

+ a **variable** is how we store information in a way that the computer can operate on
+ it has a **value:** the info itself
+ and a **name:** the cue by which we call upon the info to the computer

If you think about how variables like x and y are used in algebra, calculus, etc, that can be a useful way to conceptualize them.

Variables can contain numbers. We *assign* a value to a name.

```{r make some vars}
# We are storing the number 20 under the variable name first.var by using the left-arrow operator. It's like 20 is going "into" the label first.var
first.var <- 20
# And again with second.var
second.var <- 0.5

second.var
```

They can also contain other stuff (literally almost anything) but we'll talk about this later!

Variables live in the **environment**. The environment is your workbench in R: this is where all data are held, so that you can access and manipulate that data using R commands. It's sort of like R's working memory, where information is held for immediate use.
```{r explain what the environment is}

```

Information doesn't just exist in the environment--it can also be printed to **console.** The console is the place where you talk to R (type in commands), and R talks back to you (shows you the result of those commands).

```{r console is a special place}
# You cannot access anything that was not saved to a variable!
# Console is great for checking stuff out
# But you have to remember to save stuff to vars (into your environment) if you wanna manipulate it in any way

```

Since variables are labels for pieces of information, a variable name can be used to refer to any piece of data where you would otherwise call that data directly. For example:

```{r}
# This outputs the result to console
20 + 10
# Since first.var contains the value 20, this outputs the SAME result to console
first.var + 10
```

Variables don't just have to contain pure numbers; you can assign the output of commands to variables.

```{r}
third.var <- first.var + 10
```

You can overwrite the values of variables, by assigning something else to an existing variable name:

```{r}
# With the values of other variables
third.var <- second.var + 10
# OR of the current value of the variable. Note that R will ALWAYS use the PREVIOUS value of a variable in all the calculations BEFORE re-assigning the final result to the variable name.
third.var <- third.var + 10
# You can also delete variables from your environment
rm(third.var)
```

Variable naming has to obey certain rules.

* must **start** with a letter
* can contains the following:
    + letters (upper or lowercase, R is case sensitive!)
    + numbers
    + period . and underscore _

In addition, name your variables to maximize both informativeness and conciseness.

* Informative: `raw.data` is better than `d`, for example
    + Use a period . or underscore _ to separate words in multi-word variable names
* Concise: Consistent abbreviations are okay, e.g. `max.val` is okay instead of `maximum.value`
    + When in doubt, use lowercase letters

For example:

```{r}
# The numbers being assigned here are just dummies
period.separated.var <- 1000
underscore_separated_var <- 57
lowerCamelCaseVar <- 1.96
```

## Data types

The different formats of information R understands are called **data types.** Broadly, the type of a piece of data specifies what kind of information it holds, and what kinds of operations might be done on that data. There are different things you would want to do with different pieces of data, and data types allow a programming language to know the sorts of things you want to use a piece of information for.

### Numeric data
Literally... any (real) number

```{r}
first.var <- 280
second.var <- 41.5
```

#### Arithmetic operators

Here are (some of) the things you can do with numeric data:

```{r}
# Maths!
first.var + second.var
first.var - second.var

# All the arithmetic operators:
# + (add) - (subtract) * (multiply) / (divide) ^ (exponent)
# R obeys PEMDAS
```

```{r}
# Bonus arithmetic operator; you may need someday
10 %% 2 # The modulo %% operator returns the remainder of dividing two numbers. Observe the results for the following modulo operations
10 %% 3
10 %% 4
10 %% 5
```

### Logical data

A special case of numeric data is **logical** data, which is essentially binary (0 or 1) numeric data. R will let you represent binary numeric data as `TRUE`/`FALSE`, where 0 means `FALSE` and 1 means `TRUE`. You can do everything you can do with numeric data on logical data, like:

```{r}
TRUE + 1
FALSE + 1
```

```{r}
# But what's more interesting & useful is...
TRUE * 10 # Like multiplying by 1. the other number stays the same!
FALSE * 10 # Like multiplying by 0. the other number becomes 0!
```

Note that this is case-sensitive: R requires `TRUE` and `FALSE` to be written in all caps!

Also you can use logical data as informational flags. To be revisited...

### Character data

This is letter/word-based data. A piece of character data is called a **string.**

R understands that this data is made up of text, and treats it in a special way. You might encounter this if you are storing free-response text data in R, or perhaps if you have a list of word-based stimuli stored in R as part of your task data.

```{r}
text.var <- "statistics"
text.var.2 <- "my favorite number is 7" # Disclaimer: Not actually my favorite number
```

Here are a couple of the text-manipulating functions that specifically operate on character data:

```{r}
toupper(statistics) # You can coerce things to upper or lowercase with the right function
nchar(statistics) # You can also count the number of characters 
# Of note is that you canNOT do arithmetic on strings
not.a.number <- "7"
# Try to do arithmetic on this variable... what happens?
```

**R is not the place to be doing intense textual analysis!** But R CAN handle your run-of-the-mill text labeling/processing needs.

## Data structures
With all the data types we just talked about... just with more than one piece of data!

### Vectors

A **vector** is a sequence of pieces of information that are **all the same data type** (for the computer) and meaningfully related (for you). A vector has two features: data type (what's in the vector) and length (how many pieces of info are in the vector). You can think about a vector as one column of data.

```{r}
num.vector <- c(1, 2, 3, 4, 5)
length(num.vector)

# a single piece of data is also a vector with length 1!
length(first.var)
```

The individual pieces of data inside one vector are called **elements.**

Vectors can be any of the data types.

```{r}
char.vector <- c("apple", "banana", "cantaloupe", "dragonfruit")
length(char.vector)

logical.vector <- c(TRUE, TRUE, FALSE, TRUE)
```

As you can see, the function `c()` is the primary way to construct vectors, by listing out all the pieces of data you want to put into the vector inside the function. BUT, R provides the colon `:` operator, which creates a vector of integers from the number on the left to the number on the right.

```{r}
num.vector.2 <- 1:5
```

#### Indexing into vectors

When you need to access individual pieces of information inside of a vector, you'll do that by calling the vector and adding an additional "address" to specify which piece of data inside the vector you want. This address is called an **index.**
To tell R that you are indexing into a vector, you put the index inside hard brackets `[]` after the name of your vector variable.

##### Numeric indexing

One way to index data is to use numeric indices: these extract elements of your vector by position (the 1st element, the 18th one, etc).

You can index one piece of data:

```{r}
char.vector[1]
```

You can also index multiple pieces of data out of a vector. This returns another, shorter, vector! You do this by putting a valid numeric vector (every number in the vector must correspond to a position in the vector, so the indices can't be longer than the length of the vector) inside the brackets.

```{r}
# Remember the colon : operator from above
char.vector[1:3]
```

```{r}
# Or another valid numeric vector, using c()
char.vector[c(1, 4)]
```

Bear in mind that R starts indexing at 1; the first element in your vector is at index 1. This is different from several other languages that start indexing at 0.

You can also index to exclude elements from a vector, using negative numbers. This returns your input vector, but missing the elements you negative-indexed:

```{r}
char.vector[-2]
```


##### Logical indexing

**Logical indexing** involves indexing using hard brackets `[]` just as before, but instead of feeding in a vector of numbers corresponding to desired indices, you feed in a vector that is the _same_ length as your input vector, where every value is a piece of logical data: `TRUE` or `FALSE`.

What you're essentially doing with logical indexing is telling R "Yes index this one" or "No don't index this one" for every single value in your vector. The resultant vector you get from logical indexing is all of the values of your original vector to which you said `TRUE` ("Yes index this one").

```{r}
# Remember this vector? It should be of use here...
logical.vector
```

```{r}
# This will only work if these two vectors are the same length. Let's check:
length(logical.vector)
length(char.vector)
```

```{r}
# Which elements do you expect to be indexed here?
char.vector[logical.vector]
```

Right now, logical indexing may feel a little unwieldy relative to numeric indexing, and that's okay. Logical indexing will come in handy later!

You can do two types of operations on vectors:

#### Vectorized/parallel operations (n element input, n element output)

This takes a command and does it individually to each element in your vector, returning an analogous vector where the nth element of the output vector is the result of performing the operation on the nth element of the input vector.

If you wanna do the same thing to every element of a vector this is the way to do it fast!

```{r}
# Basically all the arithmetic operations are vectorized, for example
num.vector + 1
num.vector * 12
# and many other operations
```

#### Summarizing operations (n element input, 1 element output)

These operations require multiple pieces of data (a vector) and returns one piece of data. Stats operations are typically of this type.

```{r}
mean(num.vector)
min(num.vector)
max(num.vector)
```

### Data frames

This is the data structure where it happens! Imagine your usual rectangular Excel spreadsheet that you might have for your study data, where each column of the spreadsheet contains a meaningful category of information (e.g. subject ID, task condition, trial response), and each row contains one observation of information (one subject, or one trial within subject, etc). A dataframe is basically that.

A **dataframe** is a special rectangular data structure in R that is *column-optimized.* A dataframe is essentially a series of vectors *of equal length* stuck together in a meaningful way. 

Each column is a vector, containing all the observations of a particular meaningful grouping. Each row is the nth element across vectors, containing one complete observation from all the vectors. Each observation can be a subject, a trial, a group, whatever--any one meaningful something. In order for each row to be a complete observation, each column must be the same length so that there's a value from every column in every row.

```{r}
# This command will load a dataframe in from a study I worked in undergrad. It's saved externally as a CSV file (one that you could open in Excel if you wanted to), so you can see the commonalities between your garden-variety study data Excel spreadsheet and an R dataframe.
# Quick review question: Am I calling the below file path using an ABSOLUTE or RELATIVE path? How can you tell?
data <- read.csv(file = "../Descriptives/uncapher_2016_one_condition_dataset.csv", header = TRUE, stringsAsFactors = FALSE)
# Don't worry about the command below; I'm just shortening the dataframe. You'll understand what this means later! Just run it now.
data <- data[data$groupStatus != "", -c(1:4, 22:26)]
```

A dataframe is composed of a series of vectors; you can manipulate individual columns of a dataframe as vectors. But you can NOT manipulate individual rows of a dataframe as vectors. We will try this very soon and you'll see that an individual row of a dataframe behaves as a dataframe with only one row in it. Why might this be the case?

To access these individual columns and rows, we'll index into our dataframe. Indexing is back... with a vengeance!

#### Indexing into dataframes

Like indexing into vectors, but with more to love!
You can index whole rows, whole columns, and portions of both.

##### Indexing columns

You can index by NAME (each column in a dataframe has a NAME, like the header row in a spreadsheet), or by NUMBER (first column, second column, etc). Today, we'll learn about indexing by NAME, because this is the reproducible way to do it. Consider it this way: other people don't know what order your columns are in, but they can gather what's in a column by the name, so if someone else were reading your code, they'd be able to tell much more easily what column you were referring to.

With brackets:

This is similar to indexing a single vector using brackets, but this time, a dataframe has **two dimensions: length AND width.** This means that bracket-indexing into a dataframe expects TWO coordinates: which rows you want, *and* which columns you want. The indexing is ALWAYS `[row, column]` where the row index is the one BEFORE the comma, and the column index is AFTER. You can remember this as it's similar to (x, y) coordinates, where x goes horizontally and y goes vertically. 

```{r}
# When you want ALL rows, you leave the row index BLANK. BUT you still need the comma!
# Notice that I am calling the column name as a STRING.
data[, "subjNum"]
```

With the dollar sign:

Remember how dataframes are column-optimized? There's a special way to index columns of a dataframe that's optimized for using column names to index. Let's take a look:

```{r}
# This returns the exact same column as we got before!
# You first write the name of your dataframe, then the $, to tell R to go "inside" of that dataframe and look for the column you're about to specify.
# Then you write the column name, WITHOUT QUOTATION MARKS. NOT A STRING!
# This is why $ indexing is special: A word without quotes is a VARIABLE NAME, but a word in quotes is a STRING. These behave two very different ways in R.
data$subjNum
```

##### Indexing rows

Indexing rows works similarly to indexing columns in some ways. But, while columns of a dataframe have names, *rows of a dataframe do not have names, ONLY numerical indices.* So we CANNOT use dollar-sign indexing to index rows, ONLY bracket indexing. And we cannot use words to call rows, ONLY numbers.

```{r}
# Do we expect to call rows using the space BEFORE the comma, or AFTER the comma?
data[1, ]
# Observe that indexing a single row of a dataframe returns a dataframe with one row in it. Because R knows that each row of a dataframe is a complete observation of data, R keeps that observation in dataframe form even when there's only one observation, so that you can continue doing special dataframe stuff on it.

# Remember that you can use a VECTOR in the numerical index to index MULTIPLE indices, here calling multiple rows
data[1:4, ]
```

Now, we're going to step out of talking about dataframes directly, and we're going to talk about other data types that are useful inside of dataframes.

### Factor data

Remember character data from earlier? That's R's all-purpose data type for text-based data. But R knows that sometimes, especially in a dataframe, you might have a column composed of text to label your within-subject task conditions, between-subject groups, etc. **Factor data** is a datatype built on top of character data that gives you special properties that are useful when a column/vector contains grouping information.

```{r}
data2 <- data # Duplicating my original dataframe into a new variable, so I can keep an original copy!

# For example, here it might be useful for the column "groupStatus", which refers to a between-subjects labeling, to be in factor form.
# I'll keep working on data2 here to keep data as the raw copy.
# the function as.factor() is a COERCING function; we won't focus on these today, but these are a family of functions that take data of one type and change them into another type. Like, you could do as.numeric("9") to change it into the number 9.
data2$groupStatus <- as.factor(data2$groupStatus)
data2$groupStatus
# Observe how this column now has LEVELS; these are the different categories of the variable.
# They are organized alphabetically by default; you can reorganize the levels into a more meaningful way if you want

```

We won't worry too much more about factor data for now, but here are a couple of factor data's pros and cons relative to regular character data:

* PROS:
    + You get special properties that help when graphing
    + and also when creating models for data
* CONS:
    + factor data doesn't always behave exactly like character data
    + sometimes you don't get the outputs you expect if you were to run the same operation on character data

### Non-data data

You know what's the worst? Missing data! But it happens to the best of us. When you have a missing data point in an Excel spreadsheet, you might leave that cell blank. But in R, you need to put a placeholder in that spot. There's a special data type, `NA`, used as the missing data placeholder.

```{r}
# This dataframe column has missing data in it. When a participant didn't have any false alarms, there weren't any trials to calculate RT, for, so this data point is missing in that situation!
data2$rtFA

mean(data2$rtFA, na.rm = TRUE)
# Try running the above command after deleting the na.rm argument, aka mean(data2$rtFA)
```

# HOW

Now that you know (some of) what kinds of things R can manipulate, we will get into actually manipulating things!

## Pseudocode

It's so important!!! Be able to talk through what you want to do in plain English before you actually try to code it. This is how you will actually figure out what functions you need to do the things you want.

For example, let's say I want to do some exploration in the dataset we loaded earlier. Perhaps I want to calculate the mean hit rate for all subjects in the group "HMM". Before actually writing any code, I need to _operationalize_ my task into a series of pieces that can be directly translated into R commands.

1. First, call the column of data corresponding to HIT RATE
    1. but only the ones for "HMM" subjects
2. Then, calculate the mean of that

This one was pretty brief. Pseudocoding can be quick, but it can often be quite complex for larger tasks!

## Relational operators
Most of the time, when you want to do things in R, you want to do them in *some conditions but not others.*

**Relational operators** are the first key to making this happen. These are essentially inequality operators like the ones you would encounter in algebra.

```{r}
# The following is a RELATIONAL STATEMENT. This is a command involving a relational operator that returns TRUE or FALSE based on whether the statement is true or false.
2 > 1
# Or try this
2 < 1
# Or these
99 == 99
99 != 100
# All the relational operators:
# > (greater than) < (less than) >= (greater than or equal to) <= (less than or equal to) # == (is equal to, NOTE that it is TWO equals signs, ONE equals sign does something different)
# != (is not equal to)
# In most cases, you can use == and != to compare character data too--if you want to see if one piece of character data is the same as another piece
"word" == "word"
text.var == "statistics"
# %in% (is contained by; this is useful when you need to see whether the element on the left matches any of a vector of elements on the right)
1 %in% 1:5
"apple" %in% char.vector
"orange" %in% char.vector
# Note that %in% takes a VECTOR on the right side, and looks for FULL MATCHES from the element on the left to the elements on the right. To match a piece of character data into a single, LONGER piece of character data on the right (e.g. matching one word into a sentence), you have to use other strategies we won't get into here.
```

If we think about the pseudocode from earlier...
```{r}
# The first step of the pseudocode:
# call HIT RATE, but only the ones for HIGH CONFIDENCE
# We want to know which values of the column groupStatus are equal to "HMM"
data2$groupStatus == "HMM"
# Relational operators are vectorized--see that feeding in a vector on the left of the relational statement returns a logical vector, not just one logical!
```

Logical indexing is the best! and a frequent use case of relational statements. Back to our pseudocode--
```{r}
# Now that we see that we get a logical vector from calling data2$conf == "hi"
# Remember that a vector can be used inside bracket indexing
# And if you call a logical vector inside brackets, only the values that are TRUE will be indexed
data2$hitRate[data2$groupStatus == "HMM"]
# Now, I can complete my code! Because I have just identified all the values I'd like to calculate the mean on.
mean(data2$hitRate[data2$groupStatus == "HMM"])
```

## Logical operators
Relational statements will get you a long way. Sometimes, though, you need to know whether groups of conditions are true or false. You can combine the TRUEs and FALSEs of relational statements using Boolean logical operators.

```{r}
# These are the Boolean operators you'll need:
# ! (NOT operator; this returns the opposite of whatever you put in)
!TRUE # see what this returns
!(data2$groupStatus == "HMM") # This should return TRUE for all the times where groupStatus is "LMM"
# & (AND operator; this returns TRUE if both statements on either side are both true
1 == 1 & 2 == 2 # See what this returns
1 == 1 & 2 != 2 # Versus this
# | (Pipe, or shift-backslash--OR operator; this returns TRUE if AT LEAST ONE of the statements on either side is true)
1 == 1 | 2 == 2
1 == 1 | 2 != 2

# Relational operators also obey order of operations re: parentheses (). try the below statements
(1 == 1 & 2 != 2)
!(1 == 1 & 2 != 2)
(1 == 1) & (2 != 2)
(1 == 1) & !(2 != 2)
```

Other useful functions will also output TRUE or FALSE, so you can use these similarly to relational statements to generate logical output based on the content of some data.

```{r}
# Remember NA from before? Many times, you will want to know which elements of a dataframe column are missing.
is.na(data2$conf)
# There are many more of these is.whatever() functions!
```


## Conditional operators
Sometimes, you want to run commands or set values based on whether certain things are true or false. For example, imagine that you have some continuous outcome variable, and you want to recode that continuous outcome variable categorically.

We can use what we've learned about logical statements so far, in conjunction with some new commands!

```{r}
# Let's consider the column adhd in data2. This is the participant's score on the ADHD Adult Self-Report Scale (short form). The inventory scoring instructions say that any participant with a score of 4 or above is considered "potentially diagnostic of ADHD". If we wanted to add a variable that indicated whether someone did or did not have a suprathreshold ADHD score, what would we do?

# Pseudocode!
# First, call the column for ADHD
# Then, create a new column and fill in the following values based on the corresponding values in ADHD
# If the ADHD score is 4 or above, then label in the new column as ADHD diagnostic
# Otherwise (if the ADHD score is less than 4), label in the new column as ADHD non-diagnostic
```

Now, our pseudocode has just introduced a word we haven't previously seen: *IF.* What we need here is a command that will direct R to judge whether some condition is true, and then perform some action ONLY if that condition is true. If-based code is great for promoting reproducibility. You can just tell R to check that a certain condition is true, and you'll always get the result you want, no matter the particulars of your data.

```{r}
# The ifelse() function will allow us to knock out almost all of our pseudocode in one function
# First, ifelse() takes a TEST: this is any statement that evaluates to LOGICAL data. It's a question in R that must yield a YES or NO answer.
# then, ifelse() creates a new vector the same length as the test. each element of test is going to be filled in with some value based on whether the test came out as TRUE or FALSE for that element.
# if the condition is TRUE for an element, ifelse() fills in that spot with the second argument, which is named 'yes'
# if the condition is FALSE for an element, ifelse() fills in that spot with the third argument, which is named 'no'
# in plain english, it kind of says:
# check whether this condition is true or false
# if the condition is true, create a variable with value A
# otherwise, create a variable with value B
# ifelse() is a vectorized function! Remember that this means that if you input a vector in the first argument, you'll get a vector of the same length as your output, because the operation is done on each vector element separately.
data2$adhdCoded <- ifelse(data2$adhd >= 4, "adhd", "no adhd")
```

## Packages

Packages are bundles of functions in R, usually made by experienced programmers, that you can use to make life easier! Oftentimes, if there's something you want to do in R, there already exists a function to do it, that you can find in a package.

Since these packages don't come with base R, you have to install them.
```{r}
# You will need these packages in later parts of the workshop!
install.packages(pkgs = c("car",
                          "moments",
                          "ggplot",
                          "tidyr",
                          "dplyr",
                          "effects"))

# You can also do this through the GUI, using the "Packages" tab in the lower-right window of RStudio.
```

Then, once they're installed, you again have to tell R to LOAD the packages into your active R session. While you only have to install packages once, you have to load them every time you open R. (Why they don't load automatically? I don't know! But if you'd like to configure R to load packages automatically every time you open R, ask us after this)

```{r}
# Use library() to load a package ONCE IT'S INSTALLED.
# You can put the package name as a string
library("car")
# OR unquoted, as a variable name. Some functions don't care whether you use quotes or not, but some do.
library(moments)
library(tidyr)
library(dplyr)
library(effects)
```


## Getting assistance

+ On your own
    + The help docs!
    + The internet
    + Stack Overflow & other Q&A spaces
    + R user blogs
+ Asking someone else
    + Who to ask
    + How to ask

