---
title: "Plotting Tutorial Using ggplot!"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Return home

Click [here](https://cu-psych-r-users.github.io/cu-psych-r-tutorial) to return to the main CU psychology R tutorials page.

## 0: What is this tutorial? What's ggplot?

ggplot is a package in R that allows for highly customizable and pretty plots! Here, were going to learn a few basics of making plots using ggplot that will hopefully get you well on your way to making informative and beautiful data visualizations!


## 1: First Install/load ggplot2

```{r, warning=FALSE}
#load the package
if (!require(ggplot2)) {
  install.packages(ggplot2)
} else {
  require(ggplot2)
}
library(tidyverse)
```


## 2: Load in Sample Data (NHANES)

We're going to practice here on a dataset from the 1990 NHANES (National Health and Nutrition Examination Survey). The variables are below. 

Region  -  Geographic region in the USA: Northeast (1), Midwest (2), South (3), and West (4)

Sex     -  Biological sex: Male (1), Female (2)

Age     -  Age measured in months (we'll convert this to years below)

Urban   -  Residential population density: Metropolital Area (1), Other (2)

Weight  -  Weight in pounds

Height  -  Height in inches

BMI     -  BMI, measured in kg/(m^2)

```{r}
nhanes <- read.csv("NHANES1990.csv", stringsAsFactors = F)
nhanes$Age <- nhanes$Age/12 # convert age to years for convinience 

# Recoding factors
nhanes$Urban <- recode(nhanes$Urban, '1' = 'Metro Area', '2' = 'Non-Metro Area')
nhanes$Region <- dplyr::recode(nhanes$Region, '1' = 'Northeast', '2' = 'Midwest', '3' = 'South', '4' = 'West')
head(nhanes)
```


## 3: Scatter Plot & Basic ggplot Syntax

-`ggplot()` command usage including saving to a variable with `a <- ggplot()`

-The general format is `ggplot(data, aes(x = [x axis variable], y = [y axis variable])` 
    - x and y variables are always specified in this `aes()` subfunction

Run this line of code 

```{r}
ggplot(nhanes, aes(x = Age, y = Weight))
```
  
  * axes are set up the way we'd expect, and seem to have sensible values
  * why is nothing on this graph yet? we haven't put any graphic actually on the axes yet


**We need to tell the `ggplot()` call what kind of graphic to put on the axis**

  * A lot of the time, the syntax is `geom_[something]`


So lets do scatter with `geom_point()` first: 


```{r}
ggplot(nhanes, aes(x = Age, y = Weight)) + geom_point()

```


Wow, lots of data points! Maybe we can make the points smaller to see better

```{r}
ageWeightPlot <- ggplot(nhanes, aes(x = Age, y = Weight)) + 
  geom_point(size = .1, alpha = .3)

ageWeightPlot
```

Remember, it's a good habbit to save your plots to objects, not just draw them! 

```{r}
ageWeightPlot <- ggplot(nhanes, aes(x = Age, y = Weight)) + 
  geom_point(size = 1, alpha = .2, aes(color = factor(Region), pch = factor(Region)))

```

## 4: Dot Plot Organized by Grouping Factor

Question: what if we want to look at distribution of weights by region in the nhanes data?

With ggplot, if x is a factor (discrete, not continuous) we can plot dots as a function of the factor

```{r}
heightByUrban <- ggplot(nhanes, aes(x = Urban, y = Height)) + 
  geom_point()

heightByUrban
```

Woah! So much data, hard to see anything, let's use geom_jitter() with size = .05, width = .1 instead

```{r}

heightByUrban <- ggplot(nhanes, aes(x = Urban, y = Height)) + geom_jitter(size = .05, width = .1)
heightByUrban
```

### 4a: Visualizing Distributions with Violin Plots

Still a TON of data! Here's a great tool I really like for plotting the density of distributions of points like this!

```{r, warning=FALSE}
# We can use 'source' to pull bits of code from github, as well as local files
source("https://gist.githubusercontent.com/benmarwick/2a1bb0133ff568cbe28d/raw/fb53bd97121f7f9ce947837ef1a4c65a73bffb3f/geom_flat_violin.R") 

heightByUrban + geom_flat_violin(,
                   position = position_nudge(x = .15, y = 0), alpha = .7)
```


Now, we can really begin to se the skewedness of these weight distributions.


## 5: Summary Plots

Plotting data points is all well and good, but what if we want to use our plots to summarize distributions? We'll do that here: 

```{r}
weightByRegion <- ggplot(nhanes, aes(x = Region, y = Weight)) + stat_summary(fun.y= "mean", geom = "point")
weightByRegion


# People might want to know how to add a confidence interval
weightByRegionConf <- ggplot(nhanes, aes(x = Region, y = Weight)) + stat_summary(fun.data = "mean_cl_boot", fun.args=list(conf.int=.95))

weightByRegionConf
```

## 6: Fitting Lines To The Data

Let's say we think there might be a linear relationship between height and weight

We can use `geom_smooth` for this and `method = 'lm` specifically for a linear model
Also `level = .95` can specify confidence interval about the estimate at each x value

```{r}
heightByWeight <- ggplot(nhanes, aes(x = Height, y = Weight)) + geom_point() + stat_smooth(method = 'lm')
heightByWeight
```
Hmm, this actually looks like it's giving us some pretty bad predictions. We're not going to get into the stats of this now, but we can also plot using `auto` which is a mix of models, and might be a bit smarter



```{r}
heightByWeight <- ggplot(nhanes, aes(x = Height, y = Weight)) + geom_point() + stat_smooth(method = 'auto', level = .99)
heightByWeight
```

## 7: Titles, Axis Labels


The `labs()` command can be added to ggplot with different arguments, lke `x`, `y`, or `title` to make the plots clearer

```{r}
heightByUrban + labs(x = 'Urban or Not ', y = 'Height in Inches', title = 'Height by Urban States')
```

Note, if we want to change labels for FACTORS, not just axes, it's easier to do that using the tidyverse 

```{r}
weightByUrban <- ggplot(nhanes, aes(x = Region, y = Weight)) + geom_point() + 
  labs(x = 'Region of US', y = 'Weight (lbs)', title = 'Weight by Region')
weightByUrban
```

## 7: Facetting

Its useful to have several plots in a panel sometimes, not just one. 

So for this data set, say we want to plot relationships between height and weight, but by region

We can do this with `facet_wrap('Region')`

```{r}
facetPlot <- ggplot(nhanes, aes(x = Height, y = Weight)) + geom_point() + geom_smooth(method = 'lm') + facet_wrap('Region')
facetPlot
```

We can even do multiple factors

```{r}
#Optional to explain the scales = 'free_x' part

multiFacet <- ggplot(nhanes, aes(x = Height, y = Weight)) + geom_point() + geom_smooth(method = 'lm') + facet_wrap(c('Region', 'Urban'), scales = 'free_y')
multiFacet
```

## 8: Color

We can color either continuously or discretely, depending on how a variable is represented in r

We put `col = Height` into the `aes()` function because it's a grouping factor

Continous Example
```{r}
weightByRegion <- ggplot(nhanes, aes(x = Region, y = Weight, col = Height)) + geom_jitter(width = .1) + 
  labs(x = 'Region of US', y = 'Weight (lbs)', title = 'Weight by Region')
weightByRegion
```

Discrete Example
```{r}
discretePlot <- ggplot(nhanes, aes(x = Height, y = Weight, col = Region)) + geom_point() 
discretePlot
```


It is EASY to choose your own custom colors (as well as using R presets), but we're not going to get into that right at this moment

## 9: Themes

We can use themes to make our plots prettier, and also customize the gridlines a lot 



```{r}
discretePlot + theme_bw()
discretePlot + theme_minimal()
discretePlot+ theme_void()
discretePlot + theme_classic()

facetPlot + theme_bw() + labs(title = 'Weight by Height and Region')
```

## 10: Saving To Files

Arguments for `file` , `plot`, `dpi`, `width`, and `height`

We can use a variety of file formats

```{r, eval = FALSE}
ggsave('newPlotTest.pdf', plot = facetPlot, dpi = 300, width = 5, height = 5)
```



## 11: Final Points

Ggplot basic will get you a LONG way.

Also, there is much more ggplot can do for making your plots very pretty, and also plotting lots of complex models

Unlike excel and spss, which can often be cranky and difficult to bend to your will in customizing plots, ggplot is really easy to work with to make your graph look the way you want 
