---
title: 'Fall Tutorial: Bonus Content'
author: "Ellen Tedeschi"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 5
---

## Bonus!

During the tutorial on Wednesday, we had a couple requests for extra things people would like to cover. The main ones were writing your own function and correlations + correlation tables. These are also great topics for future R club meetings, but here's a brief overview of these two things.

```{r}
#First some set up:

if (!require(corrplot)) {
  install.packages("corrplot")
} else {
  require(corrplot)
}

if (!require(tidyverse)) {
  install.packages("tidyverse")
} else {
  require(tidyverse)
}
```

### Make your own function

This content has now been **moved** to part 2 of the programming lesson.

### Correlations

```{r}

# R has a handy build in dataset called mtcars full of nice continuous variables we can correlate. 
data("mtcars")
head(mtcars)

# Select only the variables you're interested in correlating
our_mtcars <- mtcars %>% 
  select(mpg, cyl, gear, vs)

# cor() is the basic R function for computing correlations. As arguments, it needs two vectors of the same length, or a dataframe with at least two columns. 

# Inputing two vectors produces one correlation coefficient.
cor(our_mtcars$mpg, our_mtcars$cyl)

# Inputting a dataframe of multiple columns creates a correlation matrix
cor(our_mtcars)

# However, these are rather ugly matrices. For nicer ones, I recommend the corrplot package.
# The main function is corrplot(), which takes as inputs a correlation matrix and a 'method'.

# First, save your correlation matrix
CarCor <- cor(our_mtcars)

corrplot(CarCor, method="circle")

# Instead of numbers, the matrix is now filled with colored circles. The color of the circle represents the value of the correlation, with 1 as the darkest blue and -1 the darkest red. The size and saturation of the circles also gives you information about the value of the correlation. Corrplot has many optional arguments for changing the appearance of this plot. Play around with these, and with the different methods. Type ?corrplot to see all the options.

colors <- colorRampPalette(c("black","white"))

corrplot(CarCor, method="ellipse", type="upper", 
         col=colors(50), bg="lightblue")

# to get really fancy, try corrplot.mixed instead. This lets you vizualize in multiple ways.

corrplot.mixed(CarCor)
```