---
title: "R_TUTORIAL_DESCRIPTIVES"
output: 
  html_document: 
    toc: true
    toc_float: true
--- 

# Welcome to the tutorial on statistics in R! 
Goals: 
1. Learn how to run simple statistical tests in R 
2. Learn how to visualize  the results of the models

# First things first:
let's open some packages we need for this tutorial.
```{r, include = F}
install.packages("effects")
#Effect Displays for Linear, Generalized Linear, and Other Models
#https://cran.r-project.org/web/packages/effects/effects.pdf

library(car)
library(effects)

```

# Next, let's get some data. We will use the multitasking dataset:
```{r}
# this dataset examines the relationship between multitasking and working memory.  original paper by Uncapher et al. 2016:  https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4733435/pdf/nihms712443.pdf

# we will download 2 datasets this time. 
# 1. simple wide-format dataframe, which only includes one condiiton of the task
#( therefore there is one line per subject) 
df <- read.csv("uncapher_2016_one_condition_dataset.csv")

# 2. a repeated measures version of this dataset
# which includes two conditions of Distractor Present (distPresent): 
df_rm <- read.csv("uncapher_2016_repeated_measures_dataset.csv")

# let's recode gender as factor. 
#IMPORTANT NOTE: Gender is fake data added by MVT for the purposes of this tutorial! 
df$Gender <- as.factor(df$Gender) # 1 = females, 2 = males 

# how many subjects are in this dataframe?
Nsubs <- length(unique(df_rm$subjNum))

```

#1. Correlations
Correlations are as simple as it gets in R. 
Let's test the correlation between multitasking and ADHD symptoms.
```{r}
#use the cor() function with two variables to get the correlation coefficient.

#use the cor.test() function to get the t, df, and p-value associated with your correlaton coefficient. 

```

# 2. T-tests
First, let's check that our variance is not heterogenous between the two groups. We can do this by calling the leveneTest() function, which performs the Levene's test of Equality of Variances.
```{r}
# first, let's make a dataset that excludes subjects who aren't labeled either "High" or "Low" on multitasking. We will filter out subjects whose groupStatus is blank ("").


# The levene test from the Car package to test the null that the variances are equal between groups. if p<0.05 then we would reject the null, meaning that variances are not equal (heterogenous).

# we do not reject the null (variances are equal). 

```

Next, let's run the actual t-test using the t.test() function.  
```{r}
#The first argument is the y variable (ADHD), followed by the grouping variable (Group Status).

# note:  R's default for t-test is a welch two-sample t-test, which assumes heterogenous (non-equal) variance between groups. but we know our variances are equal! 

# we can use var.equal = TRUE to conduct a "students" T-test, which assumes equal variances between groups. if you're familiar with SPSS, this is their default t-test.

```


Paired t-test: We wil use our repeated measures dataset (df_rm) to test whether dprime (hit rate/false alarm rate) varies based on within-subject condition of distractors in the working memory task (distPresent)
```{r}
# The only thing you need to change in the code for a paired t-test is add an argument: paired == TRUE.

```
#3. Anova 
A word of caution using statistics in R: there are many ways to run each statistical test. For an ANOVA, there are several didfferent packages that have different ways of running the ANOVA 'under the hood', or just use different conventions in reporting the results. To start with, we will show you how to run simple one-way and two-way anovas.

one-way anova. Does multitasking predict ADHD symptoms?

```{r}
# anova model

# anova output

 # coefficients (betas) indicating effect for each group.
```


Two way anova. Does Gender interact with multitasking (groupStatus) to predict ADHD symptoms?
```{r}
# let's use a median split to make a factor for impulsivity (high and low)

# 1. interactions the long way :


# 2. another way to write an interaction model (the short way)
# the * indicates interaction between groupStatus and bis.groups
# note that using * will automatically include the main effects of each variable in the model.

 # the results are identical to above. 

# the "effect" package allows you to quickly view the fits of your model

# note:Effect package will not show you the original data, only fits. 
# note: we recommend using ggplot, but this is helpful for quick plotting.
```

Post-hoc comparisons for 2x2 anova. Again, note there are many options for posthoc tests depending on the assumptions of your data. Here we will learn Tukey's post-hoc, which is commonly used in the ANOVA framework in psychology. 
```{r}
# We will get the post-hoc contrasts for each level of the interaction using the function TukeyHSD().
# The argument for the function is the model_name (aov_2), whch we just created above.

 # view the output.

#We can extract the 6 contrasts of the 2x2 interaction by indexing the results of the TukeyHSD() function.


 # view the output.  
```



# 4. linear regression 
Let's see whether continuous variable of multitasking predicts adhd symptoms. We will use the entire dataset here (not df.groupsonly) to examine this continuously. 
```{r}
# Linear regression follow this formula: lm( y ~ x, data) where lm() is the function of the linear model. 

# We will use the summary() function, with our model as input, to easly view a table of the results of the model.

```

The effect package provides a simple way to visualize the results of regressions too.

```{r}
# it will automatically plot the linear regression slope with a 95% confidence interval. 

# Like many R functions, there are many addtional arguments that will allow you to customize the plot and make it prettier. 
plot(effect("mmi", model1), main = "", xlab = "media multitasking", ylab ="ADHD symptoms", rug=FALSE)
```

# 5. Multiple regression using lm ()
What is the effect of mmi on impulsivity, controlling for ADHD?
```{r}
#multiple regression formula: add terms in your lm () function
# lm( y ~ x + m + z, data)

#we can see that both variables predict adhd symptoms quite strongly!
```

How do we view the results for each coefficient, controlling for the other covariates?
```{r}
# The effect package also allows you to plot the model fits for each coefficient, each controlling for all other variables in the model.


# if you only want to plot the results of one coefficient at a time, we can do that too.

```


# 6. interactions: Factors x continuous variables
First, let's visualize how gender interacts with the continuous variable of multitasking.
```{R}
#We will use the coplot() function to view the interaction relationship 
# where: y ~ x | m (moderator = Gender)


# we can add a few more arguments to make it pretty

```

Next, let's run the interaction using lm (). 

```{r}
#Similar to the anova, the * indicates an interaction, where R will include the individual coefficients for each variable automatically.


```

View the model interaction using effect plot! 
```{r}
# use effect to view the model 

# the relationship between mmi and adhd is stronger in males (gender == 2)
```

# 7. Interactions: Continuous x continuous variables 
Let's look at how impulsivity interacts with multitasking to predict ADHD symptoms.
```{r}
# now bis is the moderator 
coplot(dprime ~ mmi | bis, data =df ,number = 3,
       col = "black", bg = "blue", pch = 21) 

#let's run the model



```
Visualizing continuous x continous interactions 
```{r}
# the effect plot automatically bins the data at different levels of the moderator (bis)


#You can also just plot the lines in a single plot (without CI) using the multiline=T option. 
plot(effect("mmi:bis", model5), multiline=T, rug = F, main = "", ylab= "ADHD symptoms", xlab ="Media Multitasking")

# there are many more options in different packages, such as ggplot! 
```

# 7. more on model visualizations:
Added Variable plots (or partial regression plot) allow us to look at the fitted data of our variable of interest, controlling for all of the other stuff in your model. This plots both the fitted data and the regression line together. 
```{r}
# make the plot 

# make it prettier!
avPlot(model5, "mmi", col = "Black", col.lines = "Blue", pch = 18, lwd= 3, 
       ylab =  "ADHD symptoms", xlab ="Media Multitasking",
       main = "Partial Regression Plot: ADHD by Media Multitasking")


```

if you want to really make beautiful graphs and customizations of your models, 
we encourage you to experiment with ggplot!

This was only a brief introduction to the statistical modeling world of R. There are many more options available: mixed-effects modeling, mediations, structural equation models, bayessian estimation.
