---
title: "Statistics Data challenge"
output: html_document
---


# Data Challenge:
Now it's your turn! The datasets for the challenge are from Garcia-et-al 2015, examining regulatory mode: assessment and locomotion, in a cohort of adolescents. The original paper can be found here: https://www.ncbi.nlm.nih.gov/pubmed/?term=Garcia+D%5BAuthor%5D+2015+regulatory+mode

```{r}
df <- read.csv("Garcia-et-al-_2015_-data.csv")

```
# 1. Does Positive affect correlate with locomotion or assessment? Does negative affect correlate with locomotion or assessment? 
```{r}

```

#2. Test whether there are gender differences in locomotion or assessment
```{r}

```


#2. conduct a multiple regression to test whether Negative Affect predicts Assessment when controlling for positive affect, age and gender. Plot the results using the effect package. 
```{r}

```


# Test whether gender moderates the effects of negative affect on Assessment. Plot the interaction using the effect package.

```{r}
```


# Bonus Data Challenge
Use ggplot to visualize your results.
hint: you can add two dataframes to ggplot: raw data, and fitted data.
```{r}









```
