---
title: "R TUTORIAL: Descriptives"
output: 
  html_document: 
    toc: true
    toc_float: true
---

# Goals for this tutorial:
1. Learn how to group and categorize data in R 
2. Learn how to generate descriptive statistics in R 



# 0. Open dataset and load package
this dataset examines the relationship between multitasking and working memory. \
original paper by Uncapher et al. 2016: \ https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4733435/pdf/nihms712443.pdf

```{r}
# use read.csv function to open data into R 
df <- read.csv("uncapher_2016_repeated_measures_dataset.csv")

# we will continue using tidyverse tools
library(tidyverse)
```

# 1. Familiarize yourself with the data 
review from Data Cleaning: take a look at the basic data structure, number of rows and columns.
```{r}
# basic R functions
head(df)
nrow(df)
ncol(df)
names(df)
```


# 2. Selecting relevant variables 
Sometimes datasets have many variables that are unnecessary for a given analysis. To simplify your life, and your code, we can select only the given variables we'd like to use for now. 
```{r}
# tidyverse select() function 
df <- df %>% 
  select(subjNum, groupStatus, adhd, hitRate, faRate, dprime)

```

# 3. Basic Descriptives

## Summarizing data
Let's learn how to make simple tables of summary statistics.
First, we will calculate summary info across all data.

```{r}

 df %>% 
  summarize(min_HR = min(hitRate, na.rm = T),
            max_HR = max(hitRate, na.rm = T), 
            mean_HR = mean(hitRate, na.rm= T), 
            sd_HR = sd(hitRate, na.rm = T))

```

# 4.Grouping data
Next, we will learn how to group data based on certain variables of interest.
We will use the group_by() function in tidyverse, which will automatically group any subsequent actions called on the data. 

```{r}
 df %>% 
  # split the summary statistics by group 
  group_by(groupStatus) %>%
  summarize(meanHitRate = mean(hitRate, na.rm= T), 
            meanFalseAlarm= mean(faRate, na.rm= T),
            meanDprime = mean(dprime, na.rm = T))

```

We can group data by more than one factor. Let's say we're interested in how levels of ADHD interact with groupStatus (multitasking: high or low). We will make a factor for ADHD (mean-split), and add it as a grouping variable. 
```{r}
 df <- df %>% 
   # if adhd score is lower than mean, label "low", else label "high""
   mutate(adhdF = ifelse(adhd < mean(adhd), "Low", "High")) 
```

Then we'll check how evenly split these groups are. 
```{r}
df %>%
  # split the summary statistics by group 
  group_by(adhdF, groupStatus) %>%
  # how many data points are in each group? n() will calculate this for us.
  summarize(n = n())
```

Then we'll calculate some summary info on these groups
```{r}
df %>%
    group_by(adhdF, groupStatus) %>%
    summarize(meanHitRate = mean(hitRate, na.rm= T), 
            meanFalseAlarm= mean(faRate, na.rm= T),
            meanDprime = mean(dprime, na.rm = T))


```

# 5. Extra: Working with a long dataset 
This is a repeated measures ("long") dataset, with multiple rows per subject. This makes things a bit tricker, but we are going to show you some tools for how to work with "long" datasets.

## How many unique subjects are in the data 
```{r}
# Get a list of subjects using unique()
SubList <- unique(df$subjNum)

# how many subjects are in this dataframe?
Nsubs <- length(SubList)

```

## Make sure you know how many trials there were per subj.
```{r}
df %>%
  group_by (subjNum) %>%
  summarize (nTrials = n())

```

# 6. saving your work

## Saving tables into csv files 
```{r}
 basic_descriptives <-  df %>% 
   # if adhd score is lower than mean, label "low", else label "high""
   mutate(adhdF = ifelse(adhd < mean(adhd), "Low", "High")) %>%
  # split the summary statistics by group 
  group_by(adhdF, groupStatus) %>%
  summarize(meanHitRate = mean(hitRate, na.rm= T), 
            meanFalseAlarm= mean(faRate, na.rm= T),
            meanDprime = mean(dprime, na.rm = T))

# use write.csv() to save this out!
write.csv(basic_descriptives, file = "myDescriptives.csv")


# note : it will automatically save to your working directory unless you specify otherise 
# remember to check your working directory, use gwd() 
# to change your working directory, use setwd() 
# or, add the path manually "/dir/dir/dir/filename.csv"
```

