---
title: "R TUTORIAL: Descriptives"
output: 
  html_document: 
    toc: true
    toc_float: true
---

# Goals for this tutorial:
1. Learn basics of plotting descriptives in R 
2. Learn tools to generate descriptive statistics in R 


#install and load packages that you'll need
```{r setup, include=FALSE}


#install.packages("car")
#Companion to Applied Regression
#Functions and Datasets to Accompany J. Fox and S. Weisberg, An R Companion to Applied Regression, Second Edition, Sage, 2011.
# https://cran.r-project.org/web/packages/car/car.pdf

#install.packages("moments") 
# useful descriptive calculationss
#https://cran.r-project.org/web/packages/moments/moments.pdf


library(car)
library(moments)
library(tidyr)
library(dplyr)
```

# Let's explore a sample dataset! 
```{r}
# this dataset examines the relationship between multitasking and working memory.
# original paper by Uncapher et al. 2016:  https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4733435/pdf/nihms712443.pdf

# open the data from csv 
df <- read.csv("uncapher_2016_one_condition_dataset.csv")


```

First we will check the data to make sure 
(a) it's a dataframe
(b) what the column names are 
(c) how many rows and columns 
(d) snapshot of the data itself
```{r}
# check that it's a dataframe 


# list out column names (Variable names)

# how many rows ?


# how many columns? 

# get a snaphot of the data inside the dataframe

# how many subjects are in this dataframe?

```

# 1.A. graphing skew/ normality of variables 
plot histogram of continuous variables
```{r}
# plot histogram of media multitasking
hist(df$mmi)
# what is the skew?

# plot histogram of ADHD measure
hist(df$adhd)

# plot histogram of impulsivity measure
hist(df$bis)
```

Next, let's make the histogram prettier with some extra arguments.
base R graphics have many more options, these are just a few to get you started.
```{r}
# add light blue color using col = "light blue"
# add a yaxis label and an x axis label using ylab and xlab

hist(df$mmi, col= "light blue", main = "", ylab = "Frequency", xlab = "Media Multitasking")
```


#1.B. Quantative assessments of skew:
```{r}
#calculate skew for mmi. is it positive or negative? 

skewness(df$mmi)

#calculate kurtosis for mmi. is it positive or negative? 
kurtosis(df$mni) 
# hint: positive kurtosis means that the tails are thinner than normal distribution.

#Use the shapiro test, which tests null hypothesis that the samples came from a normal distribution.
shapiro.test(df$mmi)  
```
 

# 2. Basic Descriptives
let's learn how to get range, mean, sd, minimum, maximum, etc.
```{r}
# provide the range of mmi variable
range(df$mmi)

# average score  of mmi 
mean(df$mmi, na.rm = TRUE)

# standard deviation of mmi
sd(df$mmi, na.rm = TRUE)

# the simplest way to get summary descriptives on a variable 
summary(df$mmi)

 
# we can also combine them into a table, using the summarise function 
my_table <- summarise(df, mean=mean(mmi), stdev = sd(mmi),  min=min(mmi), max=max(mmi))

# if you want to save this table for later, write to a csv ! 
write.csv(my_table, "multitasking_summary_table.csv" )


#notes: it will automatically save to your working directory unless you specify otherise 
# remember to check your working directory, use gwd() 
# to change your working directory, use setwd() 
# or, add the path manually "/dir/dir/dir/filename.csv"
```

Quantiles 
```{r}

#The GroupStatus variable represents the 1st and 4th quartiles of media multitasking score (variable: mmi, LMM and HMM)

# Let's extract the mmi value at the 1st and 4th quartiles 
quantile(df$mmi, (0.25), na.rm = TRUE)

quantile(df$mmi, (0.75), na.rm = TRUE)

```

# 2B. Descriptives with grouping
We can also generate descriptive statistics based on grouping variables. 
Here we will look at impulsitivity (bis) by group: high or low media multitasking (HMM or LMM)

```{r}
# the group_by function is from our friend, from tidyverse.

# the first argument is your variable of interest (mmi)
# second & third arguments are your grouping variables (factors)
# group by the between subject variable: groupStatus (high, low media multitasking))
grouped <- group_by(mmi, groupStatus)

# now we will extract summary statistics from the grouped data
summarise(grouped, mean=mean(mmi), sd = sd(mmi),  min=min(mmi), max=max(mmi))

```

# 3.Boxplots 
We will use the basic boxplot function available in base R. Boxplots can be created for individual variables quite easily and give you a quick idea of the spread of your data. Here, we will make a boxplot for self-reported media multitasking (mmi).

```{r}
# make a boxplot using R base plots

# outliers are automatically labeled as dots.
# you can exclude them if you want with a simple argument 
# notice that the y axis also automatically changes!


# what if you prefer horizonal plots? 
```

#3B. Boxplots with grouping 
We can also use grouping variables in boxplots. Here we will plot ADHD symptoms by group: high or low media multitasking (HMM or LMM)
```{r}
#mmi = y variable, groupStatus is our (x) variable
# we are only interested in the high multitasking and and low multitasking groups, so we will filter the data to remove those in the middle (and therefore not labeled for groupStatus)


# let's make this one prettier by adding axis labels and colors 
boxplot(adhd ~ groupStatus, data=df.group,
  	xlab="Multitasking Group", ylab="ADHD", col = c("light blue", "light green"))
```

# 5. Relationships between two continuous variables

```{r}

# graph the relationship between ADHD (x) and impulsivity (y)

# this is the most basic plot you can make. 
#don't worry, later we will show you how to make much prettier graphs! 

```

# 6. Relationships of many variables
We can use scatterplot matrices  which are a handy & quick way of visualizing relationships between multiple variables in a dataset. This works best when you have 3-5 variables (it gets too hard to read after that).
Let's look at the relationships between multitasking, impulsitivity, and adhd symptoms.

```{r}

# We will use the pairs function 
# in order to isolate the 3 variables we want, indexing [16:18] variables 

# There are positive relationships evident between each pair of variables! 
# We will explore these relationshisp more in the statistics tutorial.
```

